A New Metropolis-Hastings Test for Fast Minibatch MCMC Inference and Optimization

### 2016-05-16 nips_2016.tex
clean the parts before the experriments

In this section, we introduce our new Metropolis-Hastings test and show how to use it to achieve
sufficient mixing and/or to reach better posterior modes. To start, we introduce two key values we
will use frequently: $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta')\prod_{i=1}^n p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^n p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}, so the terms here
consider the likelihood ratios in log space. The only difference between $\Delta$ and $\Delta'$ is
that they use $N$ and $n$ terms in the likelihood ratio, respectively. We assume $n \ll N$ (so
$\Delta'$ is substantially faster to compute) and that the $n$ data points $x_i$ represent a random
minibatch of data.

@@

For our MH test, the key is that we use a different $g$, the logistic function: $g(\Delta) =
(1+\exp(-\Delta))^{-1}$. Again, straightforward arithmetic shows that it satisfies the condition in
Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily sample from
it using the following procedure. Let $u$ be a uniform random variable $u \sim {\rm Unif}[0,1]$. At
any iteration with a current parameter $\theta_t$ and a proposed sample $\theta'$, we can compute
$\Delta$. We accept the candidate if $g(\Delta) > u$, and reject otherwise. To see why this works,
define a new random variable $X = g^{-1}(u)$ where again, $g$ is the logistic function. Notice that
$X$ has cumulative distribution function $F_X(x) = g(x)$. In other words, the CDF is precisely the
logistic function. This is because for some $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of the uniform random variable here is simply one. Thus, the criteria to accept the
candidate $\theta'$ is equivalent to whether $\Delta > X$. Since $X$ is symmetric about zero, the
acceptance criteria is equivalently expressed as $\Delta + X>0$, as shown in Figure~\ref{fig:part1}.

{\color{blue}
Daniel: The above makes sense, but I \emph{still} don't understand why we had to show that the CDF
of $X$ is the logistic function, because I never use that when I claim that our acceptance test is
now $\Delta > X$. All I need to say is this: we can pre-multiply $g^{-1}$ to $g(\Delta) \ge u$, and
our test is $\Delta > X$. Right???
}

@@

\begin{lemma}\label{lem:gaussian}
If the minibatch data is chosen randomly, then the distribution of $\Delta'$ is Gaussian.
\end{lemma}

@@

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship can be expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \epsilon, \quad \epsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\epsilon$.  The extra noise in $\Delta'$ means we can no longer perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. We can decompose
$X$ as $X = X_{\rm norm}+X_{\rm corr}$, where the first term is a zero-mean ``noise term'' which
has the same variance as $\epsilon$, and the second term is a zero-mean ``correction'' term.
These two convolve to form $X$. The criteria to accept can be rewritten as
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
This deconvolution process is visualized in Figure~\ref{fig:part2}. {\color{blue} (Daniel: note that
we will be fixing this figure, so the visualization should hopefully be a little clearer.)} We
decompose $X$ this way so that we can estimate the two components separately.  Notice that if we
increase the mini-batch size, this will cause the $X_{\rm noise}$ distribution to shrink and become
narrower.

Furthermore, since our proposed acceptance function $g(\Delta) = (1+ \exp(-\Delta))^{-1}$ satisfies
detailed balance (see Lemma~\ref{lem:detailed_balance}), if $\Delta$ is computed from the complete
data, the distribution of the sampled data will match our target distribution $f$, the posterior,
under standard MCMC assumptions (e.g., that the chain is run sufficiently long enough and mixes
well). Now consider the case of using minibatches of data. Let our true and estimated acceptance
probabilities be $P_a = g(\Delta) = \Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} >
0)$.  If our estimate of $X_{\rm norm}$ is exact, then $P_a' = P_a$, which follows by making the
inequality in Equation~\ref{eq:criteria} an equality.

@@

@@

In this section, we explore the convergence properties of our algorithm.  Define our estimate of
$X_{\rm norm}$ as $X_{\rm norm}'$, and let $X_{\xi} = X_{\rm norm} - X_{\rm norm}'$, so that
$\Delta' = \Delta + X_{\rm norm} + X_\xi$.  We also denote $F_X$ as the CDF of $X$.  Let the
transition kernel of complete data at sampling step $i$ be $T_i$, and let the transition kernel of
minibatch data at sampling step $i$ is $T_i'$.  We show that the distance of $T_i$ and $T_i'$ is
proportional to our estimate of $X_{\rm norm}$. We also define the stationary distribution obtained
by $T_i$ is $\pi$.  We show the by decreasing the step size, the distribution of our sampled data
converges to $\pi$. We use something called the \emph{total variation distance} between two kernels,
denoted as $\| \cdot \|$.

@@

Since ${\rm Var}(\Delta') < \epsilon^2 k^2 m$, we can assume the maximum value of $X_{\xi}$ is
proportional to the standard deviation of $\Delta$, i.e. $\epsilon$. Thus, we can assume
$\zeta=\epsilon C$, where $C$ is a constant factor.

\begin{theorem}
Assume the stationary distribution $\pi$ satisfies $\|T \circ P_0 - \pi\| \leq \eta \|P_0 - \pi\|$,
where $\eta \in [0, 1)$ for all distribution $P_0$. And assume there exists $\rho_t < 1$ such that
$\|T'_i(x, \cdot) - \pi_i'\| \leq 2\rho_i$. We can get $\| T_t' \circ T_{t-1}' \circ \cdots \circ T_1' P_0
- \pi_0 \| \leq \sum_{s=1}^t \{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\} \rho_s \alpha_s + \alpha_t $,
where $\alpha_i = \frac{\epsilon_i k }{1-\eta}$.
\end{theorem}

\begin{proof}
First, by Theorem 1 in~\cite{cutting_mh_2014}, $\|T \circ P_0 - \pi\| \leq \eta \|P_0
- \pi\|$ and $\|T-T'_i\| \leq \epsilon_i k$, the norm between the stationary distribution is $\|\pi
- \pi_i'\|\leq \frac{\epsilon_i k}{1-\eta}$.

Second, by Theorem 3.6 in~\cite{yang2013sequential}, we have  $\| T_t' \circ T_{t-1}' \circ \cdots
\circ T_1' P_0 - \pi_t \| \leq \sum_{s=1}^t \{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\} \rho_s
\alpha_s$, where $\alpha = \frac{\epsilon_i k}{1-\eta}$. Thus, we get:
\begin{align*}
 \| T_t' \circ \cdots \circ T_1' P_0 - \pi_0 \| &\leq \|T_t' \circ \cdots \circ T_1' P_0 - \pi_t\| + \|\pi_t - \pi_0\| \\
 &\leq \sum_{s=1}^t \left\{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\right\} \rho_s \alpha_s + \alpha_t,
\end{align*}
as desired.
\end{proof}

@@

\subsection{Mixture of Gaussians}\label{ssec:gaussians}

>>>>

In this section, we introduce our new Metropolis-Hastings test and show how to use it to achieve
sufficient mixing and/or to reach better posterior modes. To start, we introduce two key values we
will use frequently: $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') \prod_{i=1}^n p(x_i \mid \theta') q(\theta_t \mid
\theta')}{p(\theta_t) \prod_{i=1}^n p(x_i \mid \theta_t) q(\theta' \mid
\theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}, so the terms here
consider the likelihood ratios in log space. The only difference between $\Delta$ and $\Delta'$ is
that they use $N$ and $n$ terms in the likelihood ratio, respectively. We assume $n \ll N$ (so
$\Delta'$ is substantially faster to compute) and that the $n$ data points $x_i$ represent a random
minibatch of data.

@@

For our MH test, the key is that we use a different $g$, the logistic function: $g(\Delta) =
(1+\exp(-\beta \Delta))^{-1}$. Again, straightforward arithmetic shows that it satisfies the condition in
Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily sample from
it using the following procedure. At
any iteration with a current parameter $\theta_t$ and a proposed sample $\theta'$, we can compute
$\Delta$. Let $u$ be a value sampled from uniform distribution $U \sim {\rm Unif}[0,1]$. We accept the candidate if $g(\Delta) > u$, and reject otherwise. This testing process is equivalent to sample a value $x$ from distribution whose CDF is $g(x)$, and accept if $\Delta + x > 0$, reject otherwise. To see why this works,
define a new random variable $X = g^{-1}(U)$ where again, $g$ is the logistic function. Notice that
$X$ has cumulative distribution function $F_X(x) = g(x)$. In other words, the CDF is precisely the
logistic function. This is because for arbitrary $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of the uniform random variable here is simply one. Thus, the criteria to accept the
candidate $\theta'$ is equivalent to whether $\Delta > X$. Since $X$ is symmetric about zero, the
acceptance criteria is equivalently expressed as $\Delta + X>0$, as shown in Figure~\ref{fig:part1}.

{\color{blue}
Daniel: The above makes sense, but I \emph{still} don't understand why we had to show that the CDF
of $X$ is the logistic function, because I never use that when I claim that our acceptance test is
now $\Delta > X$. All I need to say is this: we can pre-multiply $g^{-1}$ to $g(\Delta) \ge u$, and
our test is $\Delta > X$. Right???
}
{\color{red}
Haoyu: If you don't show the CDF of $X$ is $g()$, how do you define the $X$. In order to use rule: $\Delta +X >0$, you need to tell people how to get $X$, thus, we need the distribution of $X$. If you just say $Delta > g^{-1}(u)$, then what's X? Then how can you decomposite the $X$ to get the $X_{corr}$? So we need to at least annouce the distribution of $X$ is $g()$.
}

@@

\begin{lemma}\label{lem:gaussian}
If the minibatch data is chosen randomly, then the distribution of $\Delta'$ is approximately Gaussian.
\end{lemma}

@@

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship can be expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The extra noise in $\Delta'$ means we can no longer perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. We can decompose
$X$ as $X = X_{\rm norm}+X_{\rm corr}$, where the first term is a zero-mean ``noise term'' which
has the same variance as $\varepsilon$, and the second term is a zero-mean ``correction'' term.
These two convolve to form $X$. The criteria to accept can be rewritten as
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
This deconvolution process is visualized in Figure~\ref{fig:part2}. {\color{blue} (Daniel: note that
we will be fixing this figure, so the visualization should hopefully be a little clearer.)} We
decompose $X$ this way so that we can estimate the two components separately.  Notice that if we
increase the mini-batch size, this will cause the $X_{\rm noise}$ distribution to shrink and become
narrower.

Furthermore, since our proposed acceptance function $g(\Delta) = (1+ \exp(-\beta \Delta))^{-1}$ satisfies
detailed balance (see Lemma~\ref{lem:detailed_balance}), if $\Delta$ is computed from the complete
data, the distribution of the sampled data will match our target distribution $f$, the posterior,
under standard MCMC assumptions (e.g., that the chain is run sufficiently long enough and mixes
well). Now consider the case of using minibatches of data. Let our true and estimated acceptance
probabilities be $P_a = g(\Delta) = \Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} >
0)$.  If our estimate of $X_{\rm norm}$ is exact, then $P_a' = P_a$, which follows by making the
inequality in Equation~\ref{eq:criteria} an equality.

@@

Our mini-batch MH test can not only be used to perform MCMC, but also can be used to obtain parameter samples around MAP/ML by adding temperature $t$ to the likelihood function. Thus, the target distribution become $f = \left ( \prod_{i=1}^N p(x_i \mid \theta') \right )^{t_i}$. We can decrease the temperature at each iteration to decline the mixing rate, so that our samples will converge to the MAP/ML point.

@@

In this section, we explore the convergence properties of our algorithm.  Define our estimate of
$X_{\rm norm}$ as $X_{\rm norm}'$, and let $X_{\xi} = X_{\rm norm} - X_{\rm norm}'$, so that
$\Delta' = \Delta + X_{\rm norm} + X_\xi$.  We also denote $F_X$ as the CDF of $X$.  Let the
transition kernel of complete data at sampling step $i$ be $T_i$, and let the transition kernel of
minibatch data at sampling step $i$ is $T_i'$.  We show that the distance of $T_i$ and $T_i'$ is
proportional to our estimate error $X_{\xi}$. We also define the stationary distribution obtained
by $T_i$ is $\pi$.  We show the by decreasing the step size, the distribution of our sampled data
converges to $\pi$. We use something called the \emph{total variation distance} between two kernels,
denoted as $\| \cdot \|$.

@@

Since ${\rm Var}(\Delta') < \epsilon^2 k^2 m $, we can assume the maximum value of $X_{\xi}$ is
proportional to the standard deviation of $\Delta$, i.e. $\epsilon$. Thus, we can assume
$\zeta=\epsilon C$, where $C$ is a constant factor.

\begin{theorem}
Assume the stationary distribution $\pi$ satisfies $\|T \circ P_0 - \pi\| \leq \eta \|P_0 - \pi\|$,
where $\eta \in [0, 1)$ for all distribution $P_0$. And assume there exists $\rho_t < 1$ such that
$\|T'_i(x, \cdot) - \pi_i'\| \leq 2\rho_i$. We can get $\| T_t' \circ T_{t-1}' \circ \cdots \circ T_1' P_0
- \pi_0 \| \leq \sum_{s=1}^t \{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\} \rho_s \alpha_s + \alpha_t $,
where $\alpha_i = \frac{\epsilon_i C }{1-\eta}$.
\end{theorem}

\begin{proof}
First, by Theorem 1 in~\cite{cutting_mh_2014}, $\|T \circ P_0 - \pi\| \leq \eta \|P_0
- \pi\|$ and $\|T-T'_i\| \leq \epsilon_i C$, the norm between the stationary distribution is $\|\pi
- \pi_i'\|\leq \frac{\epsilon_i C}{1-\eta}$.

Second, by Theorem 3.6 in~\cite{yang2013sequential}, we have  $\| T_t' \circ T_{t-1}' \circ \cdots
\circ T_1' P_0 - \pi_t \| \leq \sum_{s=1}^t \{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\} \rho_s
\alpha_s$, where $\alpha = \frac{\epsilon_i C}{1-\eta}$. Thus, we get:
\begin{align*}
 \| T_t' \circ \cdots \circ T_1' P_0 - \pi_0 \| &\leq \|T_t' \circ \cdots \circ T_1' P_0 - \pi_t\| + \|\pi_t - \pi_0\| \\
 &\leq \sum_{s=1}^t \left\{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\right\} \rho_s \alpha_s + \alpha_t,
\end{align*}
as desired.
\end{proof}

The Theorem above indicates that the difference of the target distribution and the sampled distribution have two components: $\rho_i$ which are determined by how efficient of $T'_i$, and $\alpha_i$ which are determined by error of our transition kernel. Both of these two terms are related to the step size. If step size is large, the $T'_i$ is very efficient, but the error of $T'_i$ increases. Otherwise, if step size is too small, the $T_i'$ is low efficient, but the error of $T'_i$ is low.

@@

In this section, we use three experiments to reveal the benefits of our mini-batch MH testing. 
\subsection{Mixture of Gaussians}\label{ssec:gaussians}
